From Precipitation to Prediction: Using ConvLSTM Models for Comprehensive River Runoff Forecasting

### 2024-06-07 paper/paper.qmd
Update

---
title: "From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting"
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:
  - name: Florian Börgel
    affiliations: 
      - ref: IOW
    orcid: 0000-0003-2402-304X
    email: florian.boergel@io-warnemuende.de
    corresponding: true

@@

River runoff is an important component of the global water cycle as it comprises about one third of the precipitation over land areas [@hagemannHighResolutionDischarge2020]. In the context of climate change studies, river runoff is usually generated in two ways. First, river runoff as input for ocean models can be created using hydrological models such as the Hydrological Discharge (HD) model [@hagemannHighResolutionDischarge2020]. HD models calculates the water balance using hydrological processes (e.g. snow, glaciers, soil moisture, groundwater contribution). It represents a complex forecasting tool that uses underlying physical processes. A different approach would use data-based models that integrate statistical correction, using the land surface schemes of global or regional climate models.

The relatively recent emergence of machine learning (ML) models has offered different approaches to river runoff forecasting. Accurate runoff forecasting is crucial for effective water resources management, particularly over extended periods [@fangExaminingApplicabilityDifferent2019; @tanAdaptiveMiddleLongterm2018]. Common approaches employ feed-forward artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks that have gained traction for long-term hydrological forecasting due to their excellent performance [@humphrey2016hybrid; @huang2014monthly; @ashrafi2017fully; @liu2020streamflow; @fang2022application; @kratzert2018rainfall].

@@

In this work, we demonstrate that ConvLSTM networks are a reliable method for predicting multiple rivers simultaneously, using only atmospheric forcing, even in the absence of a fully functional hydrological model with a complex parameterization. We use the Baltic Sea catchment as an example to illustrate our approach. Although the methodology we propose is universally applicable across various geographic regions, the Baltic Sea presents a challenge due to its unique hydrological characteristics, being nearly decoupled from the open ocean (see Figure). As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers. Freshwater enters the Baltic Sea through river runoff or positive net precipitation (precipitation minus evaporation) over the sea surface. The net precipitation accounts for 11 $\%$ and the river input for 89 $\%$ of the total freshwater input [@meier2002simulated]. Modelling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation. This makes the accurate modeling of river runoff especially critical for simulations pertaining to the Baltic Sea.

@@

As an alternative, we propose that this functional can be adequately represented by a combination of a ConvLSTM with a subsequent FC network. This approach eliminates the need for detailed knowledge of the involved processes and their modeling. Instead, these features can be "learned" by the network in an automated manner. Our proposed network architecture is visualized in @fig-ConvLSTM_FC and described in detail in the following sections. To provide an overview, we will discuss the main components of this architecture one-by-one.

@@

Before turning directly to the convolutional LSTM, the simpler architecture of the plain LSTM is examined which serves as a foundation for understanding the more complex ConvLSTM.

The Long Short-Term Memory (LSTM), a specialized form of Recurrent Neural Networks (RNNs), is specifically designed to model temporal sequences $\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]$ of $N_\tau$ input quantities $\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}$. This sequence is taken from a dataset given in form of a time series $\{\vec{X}^t\}$ with the endpoint coinciding with the specific element in the time series connected to time $t$, i.e. $\vec{X}^t[N_\tau] \equiv \vec{X}^t$, see @fig-ConvLSTM_FC. Here, $N_k$ represents the number of input "channels," which can correspond to different measurable quantities. The LSTM's unique design allows it to adeptly handle long-range dependencies, setting it apart from traditional RNNs in terms of accuracy (see @fig-lstm).

@@

As input the model receives 30 days of atmospheric surface fields temperature $T$, precipitation $P$, specific humidity $Q$ and wind speed $W$, with a daily resolution to predict the river runoff $R$ at the time step $\Delta t+1$ , which can be summarized as

$R_{\Delta t+1} = f\left(T_{t-30:t}, P_{t-30:t}, Q_{t-30:t}, W_{t-30:t}\right)$

with f being a function maps the 30 days of daily atmospheric surface fields data to the predicted river runoff.

The choice of atmospheric fields was based on the implemented river runoff calculation in the atmospheric model COSMO-CLM which uses these atmospheric fields to calculate an river runoff estimate.

@@

The accuracy of the model is displayed in Figure @fig-statistical-evaluationNN. As mention above for evaluation, the test dataset was utilized. The left panel panel (Figure @fig-statistical-evaluationNN a) ) illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5$\%$. The overall correlation is 0.997 with the resulting error metrics yielding a RMSE of 323.99 $m^3$/s and MAE of 249.51 $m^3$/s. While the model's performance is satisfactory, the discrepancies between the actual values and the predictions could partly be attributed to the use of a different atmospheric dataset than the one originally used to drive the E-HYPE model. However, by applying a rolling mean with a 5-day window, the prediction error is reduced to less than 1$\%$, which is acceptable for the purposes of climate modeling. Lastly, the right panel demonstrates that the distribution of residuals follows a Gaussian shape, suggesting that our model does not exhibit bias by systematically over or underestimating river runoff values.

@@

With the increasing demand of decision makers for regional climate projections, that allow to quantify regional climate change impacts, the availability of precise hydrological forecasting becomes invaluable. The quality of a projection for a coastal sea such as the Baltic Sea is too large parts based on the quality of the hyrdological conditions. In this work we analyze the implementation of Convolutional Long Short-Term Memory (ConvLSTM) networks for predicting river runoff, highlighting its potential to enhance river runoff forecasting across different coastal seas. Given the unique hydrological characteristics of the Baltic Sea, largely influenced by its limited connection to the open ocean and significant freshwater input from surrounding rivers, the region presents a critical case for the application of sophisticated forecasting models.

@@

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt. Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem. Donec molestie ex sit amet viverra egestas. Nullam justo nulla, fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

## Open research

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt. Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem. Donec molestie ex sit amet viverra egestas. Nullam justo nulla, fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

>>>>

---
title: "From Precipitation to Prediction: Using ConvLSTM Models for Comprehensive River Runoff Forecasting"
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:
  - name: Florian Börgel
    affiliations: 
      - ref: IOW
    orcid: 0000-0003-2402-304X
    email: florian.boergel@io-warnemuende.de
    corresponding: true

@@

River runoff is an important component of the global water cycle as it comprises about one third of the precipitation over land areas [@hagemannHighResolutionDischarge2020]. Therefore, accurate runoff forecasting is crucial for effective water resources management, particularly over extended periods [@fangExaminingApplicabilityDifferent2019; @tanAdaptiveMiddleLongterm2018]. In the context of climate change studies, river runoff is usually generated in two ways. First, river runoff as input for ocean models can be created using hydrological models such as the Hydrological Discharge (HD) model [@hagemannHighResolutionDischarge2020]. HD models calculate the water balance using hydrological processes (e.g. snow, glaciers, soil moisture, groundwater contribution). It represents a complex forecasting tool that uses underlying physical processes. The second approach uses data-driven models that integrate statistical correction, using land surface schemes of global or regional climate models \[QUELLE\].

With the recent rise of machine learning in climate research various model architectures have also been tested for river runoff forecasting. Common approaches employ feed-forward artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks that have gained traction for long-term hydrological forecasting due to their excellent performance [@humphrey2016hybrid; @huang2014monthly; @ashrafi2017fully; @liu2020streamflow; @fang2022application; @kratzert2018rainfall].

@@

In this work, we demonstrate that ConvLSTM networks are a reliable method for predicting multiple rivers simultaneously, using only atmospheric forcing, even in the absence of a fully functional hydrological model with a complex parameterization. We use the Baltic Sea catchment as an example to illustrate our approach. Although the methodology we propose is universally applicable across various geographic regions, the Baltic Sea represents a challenging region due to its unique hydrological characteristics, being nearly decoupled from the open ocean (see Figure). As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers. Freshwater enters the Baltic Sea through river runoff or positive net precipitation (precipitation minus evaporation) over the sea surface. The net precipitation accounts for 11 $\%$ and the river input for 89 $\%$ of the total freshwater input [@meier2002simulated]. Modelling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation. This makes the accurate modeling of river runoff especially critical for simulations pertaining to the Baltic Sea.

@@

As an alternative, we propose that this functional can be adequately represented by a combination of a convolutional Long-Short Term Memory (ConvLSTM) model with a subsequent fully connected neural network. This approach eliminates the need for detailed knowledge of the involved processes and their modeling. Instead, these features can be "learned" by the network in an automated manner. Our proposed network architecture is visualized in @fig-ConvLSTM_FC and described in detail in the following sections. To provide an overview, we will discuss the main components of this architecture one-by-one.

@@

Before turning directly to the ConvLSTM the simpler architecture of the plain Long-Short Term Memory (LSTM) model is examined which serves as a foundation for understanding the more complex ConvLSTM.

The LSTM, a specialized form of Recurrent Neural Networks (RNNs), is specifically designed to model temporal sequences $\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]$ of $N_\tau$ input quantities $\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}$. This sequence is taken from a dataset given in form of a time series $\{\vec{X}^t\}$ with the endpoint coinciding with the specific element in the time series connected to time $t$, i.e. $\vec{X}^t[N_\tau] \equiv \vec{X}^t$, see @fig-ConvLSTM_FC. Here, $N_k$ represents the number of input "channels," which can correspond to different measurable quantities. The LSTM's unique design allows it to adeptly handle long-range dependencies, setting it apart from traditional RNNs in terms of accuracy (see @fig-lstm).

@@

As input the model receives $N_{\tau}$ = 30 days of atmospheric surface fields temperature $T$, precipitation $P$, specific humidity $Q$ and wind speed $W$, with a daily resolution to predict the river runoff $\vec{R}$. The choice of atmospheric fields was based on the implemented river runoff calculation in the atmospheric model COSMO-CLM which uses these atmospheric fields to calculate an river runoff estimate.

@@

The accuracy of the model is displayed in Figure @fig-statistical-evaluationNN. As mention above for evaluation, the test dataset was utilized. The left panel panel (Figure @fig-statistical-evaluationNN a) illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5$\%$. The overall correlation is 0.997 with the resulting error metrics yielding a RMSE of 323.99 $m^3$/s and MAE of 249.51 $m^3$/s. While the model's performance is satisfactory, the discrepancies between the actual values and the predictions could partly be attributed to the use of a different atmospheric dataset than the one originally used to drive the E-HYPE model. However, by applying a rolling mean with a 5-day window, the prediction error is reduced to less than 1$\%$, which is acceptable for the purposes of climate modeling. Lastly, the right panel demonstrates that the distribution of residuals follows a Gaussian shape, suggesting that our model does not exhibit bias by systematically over or underestimating river runoff values.

@@

With the increasing demand of decision makers for regional climate projections, that allow to quantify regional climate change impacts, the availability of precise hydrological forecasting becomes invaluable. The quality of a projection for a coastal sea such as the Baltic Sea is too large parts based on the quality of the hyrdological conditions. In this work we analyze the implementation of a ConvLSTM networks for predicting river runoff, highlighting its potential to enhance river runoff forecasting across different coastal seas. Given the unique hydrological characteristics of the Baltic Sea, largely influenced by its limited connection to the open ocean and significant freshwater input from surrounding rivers, the region presents a critical case for the application of sophisticated forecasting models.

@@

The research presented in this study is part of the Baltic Earth program (Earth System Science for the Baltic Sea region, see [https://www.baltic.earth](https://www.baltic.earth/).
